---
title: "Factors Affecting Graduate Admission Decisions for Top Computer Science Programs"
author: "Superficial Intelligence (Nicholas Tanabe, Haodong Liu, Huiwen He, Xiaoyi Zhang)"
output: pdf_document
---
#Abstract
In this paper, we use data science to analyze graduate admissions data from Grad Cafe. We aim to better understand what factors are most important for graduate admissions and focus our analysis on the top 10 graduate computer science programs. We fit a logistic model with undergraduate GPA, GRE score, and student status as covariates. The model was not very accurate in prediction, with a McFadden R^2 of .000. This likely means that there are other factors that are more important to admission decisions beyond our selected covariates.

#Introduction
Graduate school admission can be a very mysterious and esoteric process. While undergraduates and prospective graduate students constantly stress about whether their GPAs or GRE scores are "good enough" to get into a top graduate program, there are many potential factors that may be important to the final decision in a holistic admission process. In this paper, we aim to use data science to better understand what factors are most important to graduate admissions. As students of data science and prospective graduate students, we will be focusing our analysis on top computer science programs. We start with exploratory data analysis to better understand the distributions and covariations of our variables of interest, and later aim to model admission probabilities for the top 10 graduate computer science programs using logistic regression.

Through our research we aim to answer several questions of interest:

* "What are the most important factors affecting graduate admissions decisions?"

* "Are international students held to a different standard in graduate admissions?"

* "Can we use data to model probabilities of acceptance?"

#The Data
```{r,include=FALSE}
#load packages, data, and clean dataset
library(knitr)
library(tidyverse)
library(car)
library(Hmisc)
library(gridExtra)
library(knitr)
library(scales)
(grad <- read_csv("data/cs_clean.csv",
    col_types = cols_only( 
      uni_name=col_character(),
      major=col_character(),
      degree=col_character(),
      season=col_character(),
      decision=col_character(),
      decision_date=col_character(),
      decision_timestamp=col_double(),
      ugrad_gpa=col_double(),
      gre_verbal=col_double(),
      gre_quant=col_double(),
      gre_writing=col_double(),
      is_new_gre=col_logical(),
      status=col_character())))
problems(grad)
```
For our analysis we will be using data on graduate admissions from GradCafe, a forum that allows students to submit their graduate school admissions decisions and details regarding their scores. A cleaned version of the dataset is provided by Debarghya Das on GitHub. The graduate school admission results database includes admission results and detailed student test scores, self-reported by prospective graduate students on https://www.thegradcafe.com/. The full dataset contains 345,303 observations and 19 variables with a mix of continuous and categorical data, but we will be limiting our analysis to the top 10 US graduate computer science programs, as ranked by US News. The dataset contains the following variables: 

1. **rowid (integer)** - An integer id of the row.

2. **uni_name (character)** - The name of the university.

3. **major(character)** - The subject of the program self-reported by students.

4. **degree (character)** - The type of degree program. The variable takes one of the following values: MS, MA, PhD, MFA, MBA, MEng, and Other.

5. **season (character)** - The season of application. The first letter indicates whether the program starts from the Fall semester or Spring semester, and then the letter is followed by the last 2 digits of the year the program starts. 

6. **decision (character)** - The admission decision. Contains five categories - Accepted, Rejected, Wait-listed, Interview and Other.  

7. **decision_method (character) ** - The method through which decision was communicated.

8. **decision_date (character)** - The date that the decision was communicated.

9. **decision_timestamp (integer)** - Timestamp of the decision.

10. **ugrad_gpa (double)** - The respondent’s undergraduate GPA. The scale of the GPA varies because some students use a 10-point scale while others use a 4-point scale.

11. **gre_verbal (double)** - GRE verbal score, which varies from 130 to 170 for the new GRE and from 200 to 800 for the old GRE.

12. **gre_quant (double)** - GRE quantitative score, which varies from 130 to 170 for the new GRE and from 200 to 800 for the old GRE.

13. **gre_writing (double)** - GRE writing score that ranges from 0 to 6.

14. **is_new_gre (logical)** - Whether or not the applicant took the new GRE.

15. **gre_subject (double)** -  GRE subject test score on a 200 to 990 score scale. 

16. **status (character)** - Status of the candidate. Can be "International", "International with US Degree", "American" or "Other".

17. **post_data (character)** - The date in which the observation was posted on grad cafe.

18. **post_timestamp (integer)** - Timestamp of the post.

19. **comments (character)** - Applicants’ comments.

We decided to drop variables which either contain little information such as 'gre_subject', which few candidates reported, and 'rowid' which is redundant, and variables which are not of interest to us, such as 'comments', 'decision_method', 'post_data', and 'post_timestamp'. It is also important to note the limitations of the data due to the self-reported nature. The dataset contains many missing values and may be biased data due to self-reporting. In addition to this, there are likely many other factors that are considered in graduate admissions such as research experience, undergraduate curriculum, and more that are not present in the data. While we tried to scrape the comments field for relevant keywords such as "research experience," the majority of observations did not include comments or relevant information in the comment field.

##Exploratory Data Analysis
First, we look at the top ten graduate programs in Computer Science as ranked by US News & World Report, which provides various rankings for US and international colleges. The full ranking can be observed at the following URL: https://www.usnews.com/best-graduate-schools/top-science-schools/computer-science-rankings 

```{r, echo=FALSE}
#Top 10 Computer Science Programs
grad1 <- grad %>% group_by(uni_name) %>% filter(decision == "Accepted") %>% count(uni_name) %>% arrange(desc(n))
grad2 <- grad %>% group_by(uni_name) %>% count(uni_name) %>% arrange(desc(n))
colnames(grad1)[2] = "accepted"
colnames(grad2)[2] = "applied"
topten <- merge(grad1,grad2,by =("uni_name")) %>% mutate(acceptance_rate = percent(accepted/applied)) %>% filter(str_detect(uni_name,"CMU")|str_detect(uni_name,"MIT")|str_detect(uni_name,"Stanford")|str_detect(uni_name,"UCB")|str_detect(uni_name,"UIUC")|str_detect(uni_name,"Cornell University")|str_detect(uni_name,"University Of Washington")|str_detect(uni_name,"GTech")|str_detect(uni_name,"Princeton")|str_detect(uni_name,"UT Austin")) %>% mutate(rank=c(1,6,8,2,9,3,4,5,10,7)) %>% arrange(rank)
topten <- topten[c(5,1,2,3,4)]
kable(topten, col.names = c("Rank",
                           "University Name",
                           "# Accepted",
                           "# Applied",
                           "Acceptance Rate"), caption="Number of Applications and Acceptance Rates for Top 10 Computer Science Programs")
top10 <- head(topten,10)$uni_name 
grad <- subset(grad, uni_name %in% top10)
```

From the table above, we can observe that there may be some outcome reporting bias ...

Next, we look at the change in number of applications over time.

```{r, fig.cap="\\label{testPlot}This figure shows the overall trend in the number of applications per year",echo=FALSE}
# Decision reported over time (2015, 2016, 2017)?
# Create a dataset for plotting number of application verses year
grad_year = grad %>% select(degree, decision_date)  %>% 
  mutate(yr = as.integer(str_c("20", decision_date %>% str_sub(-2,-1)))) %>%
  filter(degree == "MS" | degree == "PhD") %>%
  filter(as.integer(yr) < 2016 ) %>% filter(as.integer(yr) > 2005)
grad_year$decision_date <- NULL
# plot
grad_year %>% group_by(yr, degree) %>% ggplot(aes(x = as.factor(yr), fill = degree)) + geom_bar(position = "dodge")+ 
labs(x ="Year",
     y ="Count",
     title="Number of Application for Degree Type by Year")
```

The dataset has official data of report from 2006 to 2015. The application report of MS increases each year until 2015, while the application for PhD slightly drops in 2014. The overall shape has a positive trend.



Next, we plot the distribution of GRE test scores, and GPA. Because there is a variable "is_new_gre", which distinguishs between old and new GRE, we filter for only new GRE scores, as the majority of observations report new GRE scores.
```{r, fig.cap="\\label{testPlot}These figures show the distribution of student GRE quant scores, GRE verbal scores, GRE writing scores, and GPA",echo=FALSE}
#Clean Data
grad <- grad[complete.cases(grad), ] %>% filter(is_new_gre == TRUE, ugrad_gpa <=4,status!="Other")%>% mutate(decision1 = (decision=="Accepted"), gre_total = gre_verbal + gre_quant)
grad_ms <- grad %>% filter(degree=="MS")
grad_phd <- grad %>% filter(degree=="PhD")
# GRE Verbal
verbal <- grad %>% select(gre_verbal ,is_new_gre) %>% 
filter(is_new_gre == TRUE & is.na(gre_verbal)!= TRUE ) %>% ggplot + geom_histogram(aes(gre_verbal)) + 
labs(x ="GRE Verbal Score",
     y ="Count",
     title="Distribution of GRE Verbal Scores")
# GRE quant
quant <- grad %>% select(gre_quant ,is_new_gre) %>% 
  filter(is_new_gre == TRUE & is.na(gre_quant)!= TRUE ) %>% ggplot + geom_histogram(aes(gre_quant)) + 
labs(x ="GRE Quant Score",
     y ="Count",
     title="Distribution of GRE Quant Scores")
# GRE writing
writing <- grad %>% select(gre_writing ,is_new_gre) %>% 
  filter(is_new_gre == TRUE & is.na(gre_writing)!= TRUE) %>% ggplot + geom_histogram(aes(gre_writing)) + 
labs(x ="GRE Writing Score",
     y ="Count",
     title="Distribution of GRE Writing Scores")
# GPA 
gpa <- grad %>% filter(!is.na(ugrad_gpa) & ugrad_gpa < 4.0) %>% 
  ggplot(aes(ugrad_gpa)) + geom_histogram(bins = 40) + labs(titles = "Distribution of Undergrad GPA")
gridExtra::grid.arrange(verbal, quant, writing, gpa, nrow = 2, ncol = 2)
```

We see from the above histograms that GRE verbal scores range from 130 to 170 with a bell shape. Most of them concentrate 155 - 160. GRE quant score range from 130 to 170 with step like shape. Scores tend to concentrate 160 - 170. GRE writing scores range from 2 to 6 with a bell like shape. Most people get a score of 4. We also see that the distribution of GPAs for the observations tend to be left skewed, with the majority of candidates having more than 3.6 GPA. This is accepted as grad programs tend to look at GPA as a major factor, and students who aim to attend a grad school would likely have higher GPAs.

Lastly, we look at the distribution of student status (internation, US, international with US degree, etc)

```{r,fig.cap="\\label{testPlot}This figure shows the number of students for each immigration status category",echo=FALSE}
grad %>% filter(!is.na(status)) %>% 
  mutate(count = n()) %>% 
  ggplot(aes(x = status)) + geom_bar() + 
  labs(titles = "Frequency Distribution of Immigration Status") 
```

From the chart above, we see that the majority of students applying are international students. In Immigration Status, around 70% of applicants are international students and the rest of them are American and students with unclear immigration status. We can tell that a big amount of graduate or Ph.D. students are coming from an international background. 
